Model-View-Controller Package

### 2006-07-20 inst/doc/MVCClass.Rnw
Moved some of the class definitions into the BioMVCClass package so that the MVCClass package no longer depends on Biobase, graph, or Rgraphviz.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18314 bc3139a8-67e5-0310-9ffc-ced21a209358
Changed the text for the vignette, added a man page, and removed some old pictures that are not being used.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18900 bc3139a8-67e5-0310-9ffc-ced21a209358

\title{Model-View-Controller Package}
\author{Elizabeth Whalen}
\date{May 5, 2006}

@@

The purpose of the \Rpackage{MVCClass} package is to provide the definitions of
classes and generic functions that will be used by other packages to create 
model-view-controller applications.  Currently, the packages that use the 
\Rpackage{MVCClass} package are \Rpackage{iSPlot} and \Rpackage{iSNetwork}.  
Both packages use the definitions from \Rpackage{MVCClass} to create linked, 
interactive views of data.

@@

Thus, the \Robject{singleModelMVC} class binds the model, view, and controller
objects together and the \Robject{linkedModelMVC} class allows the MVC objects
to be related to each through a parent-child (tree) hierarchy because of the
slots, parentMVC and childMVCList.  These two slots also show that a MVC
object can have at most one parent MVC and can have as many children MVCs as
it wants.

@@

The model class is responsible for storing and updating the data.  These two
functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
structure for the model classes.  Here the virtual class,
\Robject{gModel}, has the slots modelData, linkData, virtualData, and
modelName.  These four slots are common information that all models will
need.  The modelData slot is the data set for this model, the linkData slot is
a list of two functions, \Rfunction{toParent} and \Rfunction{fromParent},
which link this model to its parent and child models, respectively (see Section
\ref{Sec:MVC}), the virtualData slot is data pertaining to the views
that needs to be stored with the model so that it can be shown in all views of
this model, and the modelName slot is the name of the model.

@@

For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
the modelData has class data frame (or matrix), the \Robject{graphModel} class
represents a model where the modelData has class graph, and the
\Robject{exprModel} class represents a model where the modelData has class
exprSet.

@@

@@

When creating a plot of a data set, there is a general class, called
\Robject{plotView}, that has the slots, plotDevice, plotPar, and drArea, which
store the device number of the plot, the plotting parameters, and the Gtk
drawing area object, respectively.  This class is a virtual class because it
is not intended to have any objects as it just represents a general plot.  The
plot classes that can have
objects are the specific plot classes, \Robject{sPlotView},
\Robject{graphView}, and \Robject{heatmapView}, which inherit from the
\Robject{plotView} class.

The \Robject{sPlotView} class represents a scatterplot view and it has the
extra slots, dfRows, xvar, and yvar, where dfRows are the row names from the
model that are shown in the plot, xvar is the variable name (the column name
if the model is a data frame) from the model that
is shown as the x variable in the plot, and yvar is the variable name (the
column name if the model is a data frame) from the
model that is shown as the y variable in the plot.  The \Robject{graphView}
class represents a graph plot view and it has the extra slot, grLayout, which
holds a \Robject{Ragraph} object that represents the layout of the graph plot.
The \Robject{heatmapView} represents a heatmap view of the data and it contains
the slot, ordering, which is a list that is returned from the
\Rfunction{heatmap} function to give information about the dendrogram
ordering.  Note that some of these views are only applicable for certain model
types.  For example, the \Robject{graphView} class will only make sense for a
view of a graph object, which would be stored in the \Robject{graphModel}
class.

Several generic functions have been defined that refer to operations 
performed on views.  These generic functions are \Rfunction{motionEvent}, 
\Rfunction{clickEvent}, \Rfunction{updateView}, \Rfunction{redrawView} and
\Rfunction{identifyView}.  Methods for these generic 
functions will be defined in packages that use this package (in 
\Rpackage{iSPlot} and \Rpackage{iSNetwork}).  When these methods are defined, 
the \Rfunction{motionEvent} method will be used to respond to a mouse over 
event on a view, the \Rfunction{clickEvent} method will be used to respond to 
a click event over a view, the \Rfunction{updateView} method will be used to 
update only a portion of the view, the \Rfunction{redrawView} method will 
be used to completely redraw the view, and the \Rfunction{identifyView} method
will be used to identify an object (such as a point, a node, etc.) on a view.

@@

The figure shows the inheritance structure for message objects that can 
be passed within one MVC object and a later figure in this section will show 
the inheritance structure for messages objects that can be passed between MVC 
objects.

@@

Currently, there are two types of messages for messaging within a model, view,
controller object: an add and an update message, and both of these messages
are performing some type of modification.  Thus, both the add message,
\Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
inherit from the virtual class, \Robject{gModifyMessage}, which has 
three slots, dataName, mData, and type.  dataName is the character
string that gives the name of the model, mData is a list of data needed to
perform the addition or alteration operation, and type is a character string
that gives the type of addition or alteration to perform.  Both
\Robject{gAddMessage} and \Robject{gUpdateMessage} are virtual classes.

@@

The \Robject{gUpdateDataMessage} class represents a message to update
a model.  This message is crucial for linking because it is responsible for
ensuring that once a model has been updated this information must propagate to
any views of the model and this information must also be passed along to any
other MVC object that is related to this model.  Because this information gets
passed to other MVC objects, this message also has one new slot, from, in 
addition to the slots it inherits from \Robject{gUpdateMessage}.  The from
slot is the name of the model that the update data message came from.  This
slot is necessary because the update data message can come from one of two
sources: the message can occur because of user interaction with a view
of this model or the message may occur because a different model that is
linked to this model has been updated.  So the update data message may come
from within this MVC object or it may come from a different MVC object that
has changed its model and is related to this MVC object.  
Information on passing messages between MVC
objects will be given later in this section.  Finally, the
\Robject{gUpdateViewMessage} class represents a message to update a view.

Message classes have also been defined to pass information between MVC objects.
The two main types of messages, the add and update messages, will also be used 
here.  The inheritance structure for messages passed between MVC objects is 
shown in the figure below.

@@

The first message class that will be discussed is \Robject{gAddChildMessage}.
A \Robject{gAddChildMessage} object will be very similar to a
\Robject{gAddDataMessage} object because they are both adding a new
model and a new MVC object.  The difference with a \Robject{gAddChildMessage}
object is that it must fill in some extra information to tie this new MVC
object to its parent MVC object.  The \Robject{gAddChildMessage} class
inherits from the \Robject{gAddMessage} class, with the same slots of
dataName, mData, and type.  The dataName slot is the name of the new model (and
new MVC), the mData slot is a list that contains the model data and virtual
data to fill the modelData and virtualData slots, respectively, of the new
model, and the type slot gives the type of the model (currently, the options
are ``exprSet'', ``graph'', or ``data.frame'').

@@

A generic function, \Rfunction{handleMessage}, is defined that will be used to 
read message objects.  Methods for this generic function will be defined in 
packages that use this package (\Rpackage{iSPlot} and \Rpackage{iSNetwork}).

@@

Linking callback functions to events, such as a button click, a mouse movement
or a key press event, is what allows users to have an interactive
environment.  Then when an event happens something will occur in response
depending on the callback function linked to the event.  Now to create a
flexible environment we do not want a callback function that always does the
same thing when an event occurs.  For example, when a left button press event
occurs, we may sometimes want a point to be colored and at other times we may
want a point to be hidden.  To allow this flexibility, a class called
\Robject{gEventFun} was created.  By using this class, users will be able to
change the response to an event.

The \Robject{gEventFun} class stores all of the information about a potential
callback function.  Having the callback function information stored in an 
object will allow a user to connect events to callback functions.  The class 
definition for \Robject{gEventFun} is shown in the figure below.

@@

The callback function that is stored in the callFun slot must take only one
parameter and the class of this parameter will depend on the model class in
the active MVC object.  For example, for a graph model, the
parameter will be of class \Robject{AgNode}, representing a node in the graph.
For a data frame model, the parameter will be of class character, representing
a row in the data frame.

The preprocessing functions that are stored in the preprocessFun slot will
take no parameters.  The purpose of the preprocessing functions is to set some
variables that the callback function will need.  Thus, not all callback
functions need preprocessing functions and if no preprocessing functions
are needed, then this slot is set to NULL.  If the callback function sets the
color of a node in a graph, then a preprocessing function is needed to
determine what color this will be.  So in this example, the preprocessing
function opens a color browser to allow the user to choose what the new
color is.

@@

As mentioned in Section \ref{Sec:Overview} this package is intended to be a 
behind the scenes package that is used by other packages for its class and 
generic function definitions.  Currently, this package is used by the 
\Rpackage{iSPlot} and \Rpackage{iSNetwork} packages.

>>>>

\title{Model-View-Controller Package}
\author{Elizabeth Whalen}
\date{July 19, 2006}

@@

The purpose of the \Rpackage{MVCClass} package is to provide the definitions of
classes and generic functions that will be used by other packages to create 
model-view-controller applications.  Currently, the \Rpackage{iSPlot} and the
\Rpackage{iSNetwork} packages use the \Rpackage{MVCClass} package to create
linked, interactive views of data.  The \Rpackage{BioMVCClass} package extends
the classes in the \Rpackage{MVCClass} package to include more model and view
classes.  These definitions were put in a separate R package because the
\Rpackage{BioMVCClass} packages depends on the following R packages:
\Rpackage{Biobase}, \Rpackage{graph}, and \Rpackage{Rgraphviz}.

@@

Thus, the \Robject{singleModelMVC} class binds the model, view, and controller
objects together and the \Robject{linkedModelMVC} class allows the MVC objects
to be related to each through a parent-child (tree) hierarchy because of the
slots, parentMVC and childMVCList.  These two slots also show that a MVC
object can have at most one parent MVC and can have many children MVCs.

@@

The model class is responsible for storing and updating the data.  These two
functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
structure for the model classes.  Here the virtual class,
\Robject{gModel}, has the slots: modelData, linkData, virtualData, modelName,
and modelVar.  These five slots are common information that all models need.
The modelData slot is the data set for this model, the linkData slot is
a list of two functions, \Rfunction{toParent} and \Rfunction{fromParent},
which link this model to its parent and child models, respectively (see Section
\ref{Sec:MVC}), the virtualData slot is data pertaining to the views
that needs to be stored with the model so that it can be shown in all views of
this model, the modelName slot is the name of the model (and the name of the
MVC), and the modelVar slot is a list of named model variables (for example, a
statistic for each element in the model).

@@

In this package, the only specific model class is \Robject{dfModel}.
The \Robject{dfModel} class represents a model where the modelData has class
data frame (or matrix).  Other model classes are defined in the
\Rpackage{BioMVCClass} package.

@@

The model classes may also be asked to provide information by a
\Robject{gAskAncestorMessage} object (which is discussed in Section
\ref{Sec:Message}) and thus, the generic function \Rfunction{provideInfo} has
been defined.  Again, the \Rfunction{provideInfo} methods are defined in
packages that use this package, such as \Rpackage{iSNetwork}.

@@

When creating a plot of a data set, there is a general class, called
\Robject{plotView}, that has the slots, plotDevice, plotPar, and drArea, which
store the device number of the plot, the plotting parameters, and the Gtk
drawing area object, respectively.  This class is a virtual class because it
is not intended to have any objects as it just represents a general plot.  The
plot classes that can have objects are the specific plot classes,
\Robject{sPlotView} and \Robject{qqPlotView}, which inherit from the
\Robject{plotView} class.  More view classes are defined in the
\Rpackage{BioMVCClass} package.

The \Robject{sPlotView} class represents a scatterplot view and it has the
extra slots, dfRows, xvar, and yvar, where dfRows are the row names from the
model that are shown in the plot, xvar is the variable name (the column name
if the model is a data frame) from the model that
is shown as the x variable in the plot, and yvar is the variable name (the
column name if the model is a data frame) from the
model that is shown as the y variable in the plot.  The \Robject{qqPlotView}
represents a qq-plot view and it has the extra slots, xval and yval, where
xval are the numeric values of the points on the x-axis and yval are the
numeric values of the points on the y-axis.

Several generic functions have been defined that refer to operations 
performed on views.  These generic functions are \Rfunction{motionEvent}, 
\Rfunction{clickEvent}, \Rfunction{updateView}, \Rfunction{redrawView} and
\Rfunction{identifyView}.  Methods for these generic functions are defined in
packages that use this package, such as \Rpackage{iSPlot} and
\Rpackage{iSNetwork}.  When these methods are defined, the
\Rfunction{motionEvent} method responds to a mouse over event on a view, the
\Rfunction{clickEvent} method responds to a click event over a view, the
\Rfunction{updateView} method updates only a portion of the view, the
\Rfunction{redrawView} method completely redraws the view, and the
\Rfunction{identifyView} method identifies an object (such as a point, a node,
etc.) on a view.

@@

The figure shows the inheritance structure for message objects that can 
be passed within one MVC object and Figure \ref{Fig:BetwMess} in this section
shows the inheritance structure for messages objects that can be passed
between MVC objects.

@@

Currently, there are two types of messages for messaging within a model, view,
controller object: an add and an update message, and both of these messages
are performing some type of modification.  Thus, both the add message,
\Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
inherit from the virtual class, \Robject{gModifyMessage}, which has 
three slots, dataName, mData, and type.  dataName is the character
string that gives the name of the model, mData is a list of data needed to
perform the addition or alteration operation, and type is a character string
that gives the type of addition or alteration to perform.  Both
\Robject{gAddMessage} and \Robject{gUpdateMessage} are also virtual classes.

@@

The \Robject{gUpdateDataMessage} class represents a message to update
a model.  This message is crucial for linking because it is responsible for
ensuring that once a model has been updated this information must propagate to
any views of the model and this information must also be passed along to any
other MVC object that is related to this model.  Because this information gets
passed to other MVC objects, this message also has one new slot, from, in 
addition to the slots it inherits from \Robject{gUpdateMessage}.  The from
slot is the name of the model that the update data message came from.  This
slot is necessary because the update data message can come from one of two
sources: the message can occur because of user interaction with a view
of this model or the message may occur because a different model that is
linked to this model has been updated.  So the update data message may come
from within this MVC object or it may come from a different MVC object that
has changed its model and is related to this MVC object.  Information on
passing messages between MVC objects is given later in this section.  Finally,
the \Robject{gUpdateViewMessage} class represents a message to update a view.

Message classes have also been defined to pass information between MVC
objects.  The message types for passing information between \Robject{MVCs} are
similar to the messages that are passed within \Robject{MVC} objects.  Now the
add child message is similar to the add model message, and the send parent
message and the send child message are similar to the update model message.
The inheritance structure for messages passed between MVC objects is shown in
the Figure \ref{Fig:BetwMess}.

@@

The first message class that will be discussed is \Robject{gAddChildMessage}.
A \Robject{gAddChildMessage} object is very similar to a
\Robject{gAddDataMessage} object because they are both adding a new
model and a new MVC object.  The difference with a \Robject{gAddChildMessage}
object is that it must fill in some extra information to tie this new MVC
object to its parent MVC object.  The \Robject{gAddChildMessage} class
inherits from the \Robject{gAddMessage} class, with the same slots of
dataName, mData, and type.  The dataName slot is the name of the new model (and
new MVC), the mData slot is a list that contains the model data and virtual
data to fill the modelData and virtualData slots, respectively, of the new
model, and the type slot gives the type of the model.

@@

Finally, the \Robject{gAskAncestorMessage} class, which inherits from the
\Robject{gMessage} class, is used to to ask an ancestor MVC for information
about its model.  This class is used when a child MVC is being created and it
needs information from more than just its immediate parent, i.e. it also needs
some information from its grandparent or a different ancestor.  The
\Robject{gAskAncestorMessage} class has three slots, from, type, and mData.
The from slot is the name of the MVC that asked for the information (i.e. where
this message came from), the type slot is the type of model to look for in an
ancestor, and the mData slot is the information being asked for in the model.

A generic function, \Rfunction{handleMessage}, is defined that is used to 
read message objects.  Methods for this generic function are defined in 
packages that use this package, such as \Rpackage{iSPlot} and
\Rpackage{iSNetwork}.

@@

Linking callback functions to events, such as a button click, a mouse movement
or a key press event, is what allows users to have an interactive
environment.  Then when an event happens something occurs in response
depending on the callback function linked to the event.  Now to create a
flexible environment we do not want a callback function that always does the
same thing when an event occurs.  For example, when a left button press event
occurs, we may sometimes want a point to be colored and at other times we may
want a point to be hidden.  To allow this flexibility, a class called
\Robject{gEventFun} was created.  By using this class, users are able to
change the response to an event.

The \Robject{gEventFun} class stores all of the information about a potential
callback function.  Having the callback function information stored in an 
object allows a user to connect events to callback functions.  The class 
definition for \Robject{gEventFun} is shown in Figure \ref{Fig:EventFun}.

@@

The callback function that is stored in the callFun slot must take only one
parameter and the class of this parameter depends on the model class in
the active MVC object.  For example, for a data frame model, the parameter
is of class character, representing a row in the data frame.

The preprocessing functions that are stored in the preprocessFun slot
take no parameters.  The purpose of the preprocessing functions is to set some
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions and if no preprocessing functions
are needed, then this slot is set to NULL.  If the callback function sets the
color of a node in a graph, then a preprocessing function is needed to
determine what color is used.  So in this example, the preprocessing
function opens a color browser to allow the user to choose what the new
color is.

@@

As mentioned in Section \ref{Sec:Overview} this package is intended to be a 
behind the scenes package that is used by other packages for its class and 
generic function definitions.  Currently, this package is used by the 
\Rpackage{iSPlot} and \Rpackage{iSNetwork} packages, and is extended by the
\Rpackage{BioMVCClass} package.
